NoTCP: steganographic tricks to bypass middleboxes

### 2014-07-16 paper/paper.tex
first submitted version

TCP is everywhere in the Internet and it carries the overwhelming majority of bytes~\cite{John:2007dv}. TCP is used for all types of traffic from bulk download to video streaming to instant messaging. Over the last decade a number of new protocols have been proposed to replace TCP. However many middleboxes will actively block or interfere with traffic that does not conform to their view of what TCP should look like. Consequently, few of these protocols have found any leverage.

One of the main requirements to extend and modify TCP is the ability to exchange control information. Traditionally this is what TCP option space was used for. However option space is limited, heavily contended~\cite{Handley:vj,Mazieres:uz} and network middleboxes make it difficult to use~\cite{Honda,Guha2005tb,UntoldMiddlebox2011}. On the other hand, modern networks demand new protocols for mobile and multi-homed devices~\cite{Mazieres:uz}, ubiquitous traffic encryption~\cite{Handley:vj,Rescorla:R080FoGB}, energy efficiency~\cite{Aucinas:2013uk}, connection parameters tuning for specific workloads and environments~\cite{Dukkipati:2010hs,Erman:2013uz} that were unknown 30 years ago.

We believe that the solution is to work round the middleboxes by using steganographic techniques~\cite{Frczek:2012dl,Zielinska:2014fn,Rowland:1997vq,Murdoch:2005fz} to conceal control information within the TCP header. This could be as simple as using one of the currently undefined flags to toggle the meaning of existing fields, through to using the urgent pointer without setting the urgent flag or even hashing information into the initial sequence number in much the way that SYN Cookies~\cite{Eddy:2007to} do currently.

This paper makes three main contributions. 
\begin{enumerate}
\item We discuss the possible ways to conceal information in the TCP header and give some examples of the extensions this could enable
\item We describe a measurement application that we have written to run on Android phones for a large-scale study.  
\item We present the results of an initial small scale deployment of our application.
\end{enumerate}

We use our initial study to design a way of exchanging a small amount of control information successfully across all of our tested networks and show a potential mechanism to use it for bootstrapping a full control stream currently missing from TCP. We hope that our successful results will motivate the research community into exploring the full range of possibilities in this space.

@@

\begin{figure}[t!]
\centering
\small{
\includegraphics[width=\columnwidth]{figs/tcp-header}
% \vspace{-5mm}
\caption{TCP Header with overloaded fields and extra control information (Checksum Correction / Control Stream) in payload}
\label{fig:header}
}
% \vspace{-4mm}
\end{figure}

Figure~\ref{fig:header} shows the TCP header with fields that we propose for reuse (defined later in the paper). The structure of the TCP header is very simple and gives few choices for hiding extra control information:
\begin{itemize}
    \item Setting header fields without the corresponding flags so they have no meaning to legacy implementations.
    \item Assigning one or more of the reserved field bits.
    \item Overloading the semantics for valid header fields.
\end{itemize}

@@

Only two header fields have their semantics set by corresponding flags. These are the \emph{Acknowledgment Number} and \emph{Urgent Pointer} fields. The former must also always be set after the initial SYN packet.

The general recommendation is to not use the urgent mechanism~\cite{Gont:2011vi}, therefore it could be reused for other purposes if both endpoints understand the new semantics. On the other hand, Network Intrusion Detection Systems (NIDS) often reset the field as it makes it difficult to track application-layer data~\cite{seolma}. Importantly, the field becomes \emph{Non-Urgent}~\cite{Kuhlewind:2014vd} when URG is not set -- it has no meaning for legacy implementations and so can be redefined.

The Acknowledgment Number field is only available in the initial SYN packet. It must therefore be combined with another mechanism to acknowledge control information unless TCP Simultaneous Open is used. Simultaneous open is still viable in most cellular networks~\cite{UntoldMiddlebox2011}, but we also explore other options.

@@

Three TCP Reserved bits remain unallocated. One further bit is only allocated to the experimental ECN nonce\cite{Ely:uc}. Despite their limited size, there are proposals that make use of new reserved bits due to implementation simplicity and non-interference with legacy implementations. One suggestion for reducing Web latency~\cite{Flach:2013uy} by mitigating the effects of a small number of lost packets for short connections is to send multiple copies of packets. The proposal deliberately sends copies, but uses a reserved flag to explicitly mark solution avoids triggering fast retransmit by such duplicate acknowledgments by marking them as duplicates in a reserved flag.

@@

There are multiple benefits to using the checksum field. Firstly, it is only used for error detection. It may be unnecessary when reliable error detection is done at the layer below, however link-level checks have been shown to be unreliable~\cite{Stone:2000fc}. Instead, it may be replaced with potentially stronger mechanisms \eg MPTCP DSS checksum and TCPCrypt Message Authentication Code or left for the upper layer. Secondly, checksum operations are efficient as they only involve one's complement addition. We show later in the paper (\S\ref{sec:protocol}) how checksums can be generated in a way that the intended value is recovered by the receiver after traversing multiple layers of Network Address Translation (NAT).

@@

To differentiate between general behavior of particular network and application-specific optimizations (\eg HTTP acceleration) we repeat our tests on a few application-specific ports as well as a random one:
\begin{itemize}
    \item 80, 443 - HTTP and HTTPS
    \item 993 - Secure IMAP port
    \item 5228 - Google cloud messaging port
    \item 6969 - random port number
    \item 8000 - common HTTP proxy port number
\end{itemize}

@@

We confirmed the presence of a ByteMobile proxy on Cellular 2. It replaces the original connections with new ones, adding options not present in original packets, changing window size and resetting other fields to different values to optimize TCP for cellular networks~\cite{Ha:2006td}. Such behavior makes deployment of any protocol modifications or extensions difficult -- we verified that TCPCrypt fails on this network. We do not currently have an estimate of how widespread ByteMobile proxies are as we only observed them on one network.

@@

Proxies potentially change every aspect of the original connection, rewrite or filter header values they do not recognize so hiding any information within the packet becomes difficult. We tested for them on a number of popular port numbers:
21, 22, 25, 80, 110, 135, 139, 143, 161, 443, 445, 465, 585, 587, 993, 995, 1194, 1723, 5060, 6881, 9001.

@@

We use the embedded opcodes in the handshake to exchange sufficient control information to bootstrap a complete control stream. After the handshake a higher-bandwidth \emph{Control Channel} can be defined by using TLV (Type-Length-Value) encoding to delimit control information within packet payload~\cite{Bonaventure:wx}. The channel is bootstrapped using an extended three-way handshake with fields shown in Figure~\ref{fig:header}:

\begin{enumerate}
\item \textbf{SYN}: the active opener sends SYN packet with \emph{$Opcode Namespace$ + $Opcode$} and ACK unset.
\item \textbf{SYN-ACK}: if passive opener recognizes the namespace and the $Opcode$, it replies with SYN-ACK with $Opcode_U$ = $Opcode$ (URG unset) to acknowledge the Opcode.
\item If passive opener does not receive an ACK, retransmit with $Opcode_C$ and \emph{Checksum Correction}. If that fails as well, retransmit the SYN-ACK with an invalid checksum. Finally, fall back to vanilla TCP and retransmit a standard SYN-ACK.
\item \textbf{ACK}: when active opener receives a packet, it checks for $Opcode_U$ (if URG is unset). Otherwise, it checks if Checksum = $Opcode_C$ as described above. If no Opcode is recoverable, fall back to vanilla TCP and transmit a standard ACK.
\item Otherwise both openers recognize the $Opcode$. Active opener sends an ACK packet with TLV-encoded control information in packet payload. The control stream has been successfully bootstrapped.
\end{enumerate}

@@

Figure~\ref{fig:handshake} shows an example handshake. In this case a client (active opener) sends a SYN packet as show above. The packet traverses a middlebox that does not remove the opcode and reaches the Server (passive opener). The Server replies with SYN-ACK with U\_Opcode, but the middlebox discards it. The server then retransmits SYN-ACK with $Opcode_C$ and Checksum Correction bits, delivered to the Client. The client recovers and verifies the received opcode. Hence both endpoints understand the meaning of the opcode, therefore the client replies with an ACK with control information acknowledging reception of the $Opcode$ from the Server.

@@

TCP implements an \emph{urgent mechanism} that allows the sender to make the receiver accept some \emph{urgent data}. The mechanism is often quoted as providing ``out-of-bound'' data delivery and may be compelling to use for control information even though it is explicitly not designed as such. There are ambiguities regarding the semantics of the urgent pointer and Network Intrusion Detection Systems (NIDS) tend to clear the URG flag and pointer. Hence the general recommendation is against the use of the mechanism~\cite{Gont:2011vi}.

@@

In the long term we believe that each protocol-related standard should explicitly define how it affects header fields. Examples include implicitly extending sequence number space~\cite{Mazieres:uz}, allowing encrypted TCP session to omit the timestamp option, using stronger message integrity checks making checksum field redundant, changing semantics of duplicate ACKs~\cite{Handley:vj,Flach:2013uy} or even simple window size modifications~\cite{Crowcroft:1998dk}.

We hope to map such header field modifications by existing extensions as well as network middlebox behavior patterns into an algebraic form as we have done in the simple case of checksum. This way we would not only be able to exchange control information across the network reliably, but also identify network path behavior by looking at its performed operations. Thus extending a protocol would become a matter of defining a new set of such header algebra rules hence simplifying the process of design and testing.

More importantly, this paper hopes to motivate the research community to explore the full range of possibilities in this space as well as contribute to our work by downloading our research app and helping to build a comprehensive database of middlebox interference with protocols across different networks.

>>>>

Over the last decade a number of new protocols have been proposed to replace TCP. However many middleboxes will actively block or interfere with traffic that does not conform to their view of what TCP should look like~\cite{Handley:2001vp,Vutukuru:2008fc,Guha2005tb}. Consequently, few of these protocols have seen real deployment.

One of the main requirements to extend and modify TCP is the ability to exchange control information. Traditionally this is what TCP option space was used for. However option space is limited, heavily contended~\cite{Handley:vj,Mazieres:uz} and network middleboxes make it difficult to use because of inconsistent interpretation of the specifications~\cite{Honda,Guha2005tb,UntoldMiddlebox2011}. On the other hand, modern networks demand new protocols for mobile and multi-homed devices~\cite{Mazieres:uz}, ubiquitous traffic encryption~\cite{Handley:vj,Rescorla:R080FoGB}, energy efficiency~\cite{Aucinas:2013uk}, connection parameters tuning for specific workloads and environments~\cite{Dukkipati:2010hs,Erman:2013uz} that were unknown 30 years ago.

One solution is to work round the middleboxes by using steganographic techniques~\cite{Frczek:2012dl,Zielinska:2014fn,Rowland:1997vq,Murdoch:2005fz} to conceal control information within the TCP header beyond the sight of middleboxes. This could be as simple as using one of the currently undefined flags to toggle the meaning of existing fields, through to using the urgent pointer without setting the urgent flag or even hashing information into the initial sequence number in much the way that TCP Fast Open~\cite{Chu:2011tn} does for application data delivery.

Steganographic communication channels are typically low bandwidth and statistical analysis of a sufficient amount of data usually exposes the existence of such channels~\cite{Murdoch:2005fz}. Nevertheless, it is the perfect choice for exchanging a small amount of end-to-end control data in real-time and it does not compromise network security since the channel can be detected using existing techniques. In our work we discuss the possible ways to conceal information in the TCP header and present the results of a preliminary study on the suitability of various header fields for the purpose.

We use our initial study to design a way of exchanging a small amount of control information successfully across all of our tested networks and show a potential mechanism to use it for bootstrapping a full control stream currently missing from TCP.

Our approach is not a short-term hack. Many recent proposals have demonstrated that control over end-to-end protocol behavior should remain at the endpoints. These include TCPCrypt which offers improved communication confidentiality and security and TCP Fast Open for better communication performance. We hope that our successful results will motivate the research community into exploring the full range of possibilities in this space.

@@

\begin{figure}[t!]
\centering
\small{
\includegraphics[width=\columnwidth]{figs/tcp-header}
% \vspace{-5mm}
\caption{TCP Header with overloaded fields and extra control information (Checksum Correction / Control Stream) in payload. Fields in gray used for control information. Light-gray fields are optional and are within payload.}
\label{fig:header}
}
\vspace{-4mm}
\end{figure}

Design of a control channel is constrained by protocol specification and its interpretation by middleboxes. In the past, TCP sequence numbers~\cite{Murdoch:2005fz}, timestamps~\cite{Giffin:2002wh} and IP ID~\cite{Ahsan:2002wc} have been used for encoding hidden information. In our work we focus on a different set of TCP header fields (Fig.~\ref{fig:header}, in gray, defined in \S\ref{sec:protocol}). Overall, the header leaves few options for hiding control information:
\begin{itemize}
    \item Setting fields without the corresponding flags so they have no meaning to legacy implementations.
    \item Assigning one or more of the reserved field bits.
    \item Overloading the semantics for valid header fields.
\end{itemize}

@@

Only two header fields have their semantics set by corresponding flags. These are the \emph{Acknowledgment Number} (which must always be set after the initial SYN packet) and the \emph{Urgent Pointer}.

The Acknowledgment Number field is only available in the initial SYN packet, but is usable across our tested networks (\S\ref{sec:network}). It must therefore be combined with another mechanism to acknowledge control information unless TCP Simultaneous Open is used. Simultaneous open is still viable in most cellular networks~\cite{UntoldMiddlebox2011}, but we also explore other options.

The general recommendation is to not use the urgent mechanism~\cite{Gont:2011vi}. The \emph{Urgent Pointer} could therefore be reused for other purposes as long as both endpoints understand the new semantics and middleboxes permit it. On the other hand, Network Intrusion Detection Systems (NIDS) often reset the field as it makes it difficult to track application-layer data~\cite{seolma}. Importantly, the field becomes \emph{Non-Urgent}~\cite{Kuhlewind:2014vd} when URG is not set -- it has no meaning for legacy implementations and so can be redefined if middleboxes permit it as they often do (\S\ref{sec:network}).

@@

Three TCP reserved bits remain unallocated. One further bit is only allocated to the experimental ECN nonce\cite{Ely:uc}. Despite their limited size, there are proposals that make use of new reserved bits due to implementation simplicity and non-interference with legacy implementations. One suggestion for reducing Web latency~\cite{Flach:2013uy} by mitigating the effects of a small number of lost packets for short connections is to send multiple copies of packets. The proposal deliberately sends copies, but uses a reserved flag to explicitly mark solution avoids triggering fast retransmit by such duplicate acknowledgments by marking them as duplicates in a reserved flag.

@@

There are multiple benefits to using the checksum field. Firstly, it is only used for error detection. It may be unnecessary when reliable error detection is done at the layer below, however link-level checks have been shown to be unreliable~\cite{Stone:2000fc}. Instead, it may be replaced with potentially stronger mechanisms such as MPTCP DSS checksum and TCPCrypt Message Authentication Code or left for the upper layer. Secondly, checksum operations are efficient as they only involve one's complement addition. We show later in the paper (\S\ref{sec:protocol}) how checksums can be generated in a way that the intended value is recovered by the receiver after traversing multiple layers of Network Address Translation (NAT).

@@

To differentiate between general behavior of particular network and application-specific optimizations (\eg HTTP acceleration) we repeat our tests on a few application ports as well as a random one:
\begin{itemize}
    \item 80, 443 - HTTP and HTTPS
    \item 993 - Secure IMAP port
    \item 5228 - Google cloud messaging port
    \item 6969 - random port number
    \item 8000 - common HTTP proxy port number
\end{itemize}

@@

We confirmed the presence of a ByteMobile proxy on Cellular 2. It replaces the original connections with new ones, adding options not present in original packets, changing window size and resetting other fields to different values to optimize TCP for cellular networks~\cite{Ha:2006td}. Such behavior makes deployment of protocol extensions difficult -- we verified that TCPCrypt fails on this network. We do not currently have an estimate of how widespread ByteMobile proxies are as we only observed them on one network.

@@

Proxies potentially change every aspect of the original connection, rewrite or filter header values they do not recognize so hiding any information within the packet becomes difficult.

We tested for them on a number of popular port numbers, from those use for HTTP and mail to typically encrypted ones:
21, 22, 25, 80, 110, 135, 139, 143, 161, 443, 445, 465, 585, 587, 993, 995, 1194, 1723, 5060, 6881, 9001.

@@

Covert channel is only needed to ensure that both endpoints understand the same protocol semantics. Once the handshake with hidden control information is complete, any aspect of a protocol can be redefined. For example, a higher-bandwidth \emph{Control Channel} can use TLV (Type-Length-Value) encoding to delimit control information within packet payload~\cite{Bonaventure:wx}. The channel is bootstrapped using an extended three-way handshake with fields shown in Figure~\ref{fig:header}:

\begin{enumerate}
\item \textbf{SYN}: the active opener (client) sends SYN packet with \emph{$Opcode Namespace$ + $Opcode$} and ACK unset.
\item \textbf{SYN-ACK}: if passive opener (server) recognizes the  $Namespace$ and the $Opcode$, it replies with SYN-ACK with $Opcode_U$ = $Opcode$ (URG unset) to acknowledge the Opcode.
\item If the server does not receive an ACK, retransmit with $Opcode_C$ and \emph{Checksum Correction}. If that fails as well, retransmit the SYN-ACK with an invalid checksum. Finally, fall back to vanilla TCP and retransmit a standard SYN-ACK.
\item \textbf{ACK}: when the client receives a packet, it checks for $Opcode_U$ (if URG is unset). Otherwise, it checks if Checksum = $Opcode_C$ as described above. If no Opcode is recoverable, fall back to vanilla TCP and transmit a standard ACK.
\item Otherwise both openers recognize the $Opcode$. Active opener sends an ACK packet with control information in the field it received it on and potentially TLV-encoded control information in packet payload. The control stream has been successfully bootstrapped.
\end{enumerate}

@@

Figure~\ref{fig:handshake} shows an example handshake. In this case a client sends a SYN packet as described above. The packet traverses a middlebox that does not remove the opcode and reaches the server. The server replies with SYN-ACK with $Opcode_U$, but the middlebox discards it. The server then retransmits SYN-ACK with $Opcode_C$ and Checksum Correction bits, delivered to the client. The client recovers and verifies the received opcode. Hence both endpoints understand the meaning of the opcode, therefore the client replies with an ACK with control information acknowledging $Opcode$ reception.

@@

TCP implements an \emph{urgent mechanism} that allows the sender to make the receiver accept some \emph{urgent data}. The mechanism is often quoted as providing ``out-of-bound'' data delivery and may be compelling to use for control information even though it is explicitly not designed as such. There are ambiguities regarding the semantics of the urgent pointer and Network Intrusion Detection Systems (NIDS) tend to clear the URG flag and pointer. Hence the general recommendation is not to use of the mechanism~\cite{Gont:2011vi}.

@@

In the long term each protocol-related standard should explicitly define how it affects header fields. Examples include implicitly extending sequence number space~\cite{Mazieres:uz}, allowing encrypted TCP session to omit the timestamp option, using stronger message integrity checks making checksum field redundant, changing semantics of duplicate ACKs~\cite{Handley:vj,Flach:2013uy} or even simple window size modifications~\cite{Crowcroft:1998dk}.

We hope to map such header field modifications by existing extensions as well as network middlebox behavior patterns into an algebraic form as we have done in the simple case of checksum. Such definitions would allow us to go beyond the simple opcode exchange and signal detailed protocol information between endpoints using a low-bandwidth channel. Thus extending a protocol would become a matter of defining a new set of such header algebra rules hence simplifying the process of design and testing.

We have demonstrated the feasibility of using steganography to exchange end-to-end control information across current networks. Importantly, this is not a short term solution: TCP Fast Open is already being deployed and offers even more opportunities to hide information in the initial packets within higher-level protocol information, such as HTTP headers and cipher suites~\cite{Zielinska:2014fn}. Even if TCP was completely replaced by another protocol, opportunities for control information exchange remain.
